Grow with the flow: faster growth is associated with more variable precipitation in a perennial herb

### 2016-02-02 ms/ms.Rnw
Edited manuscript.

% Local adaptation and clines
Local adaptation within species is ubiquitous; populations generally have higher fitness in their native environment, but perform poorly outside it \citep{Schluter_2000, Hereford_2009}. Local adaptation also frequently leads to clines in both phenotypes and allele frequencies when selection varies over environmental gradients \citep{Huxley_1938, Endler_1977}. Phenotypic differences between populations along the cline most often have a genetic basis and can be studied in a common garden \citep{Turesson_1922, Clausen_etal_1940, Hiesey_etal_1942}. Despite a long history of studying local adaptation and clines, it remains to challegenging to identify exactly which traits are under selection and which differ for nonadaptive reasons. In particular, the role that physiological differences play in local adaptation is poorly understood, despite the fact that physiology is frequently assumed to explain adaptation to the abiotic environment. We need to understand physiological adaptations within species as a baseline for anticipating how organisms will respond to climate change. A related problem is identifying which features of the environment, abiotic factors like soil water availability or biotic interactions, cause spatially varying selective pressures. Here, we examine physiological trait variation and possible selective agents in the perennail herb \textit{Mimulus cardinalis} Douglas ex Benth. (Phrymaceae), a model system for local adaptation studies.

% Intrinsic versus plastic variation
There are two basic approaches one can use to identify candidate traits underlying local adaptation in a common garden. First, if genetically-based trait differences between populations vary clinally with environmental differences, this may point to traits important for local adaptation. Second, genotype by environment interactions could indicate that variation in plasticity mediates local adaptation. We distinguish between these signatures of local adaptation by referring to `intrinsic' and `plastic' trait variation, respectively. There are classic cases adaptation involving both intrinsic and plastic trait variation. For example, intrinsic differences in critical photoperiod [CITE] and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment. Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genotype by environment interactions cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}.

% Selective agents
Either instrinsic and/or plastic variation should vary clinally along environmental gradients. Indeed, clines in ecologically important traits are widespread in nature \citep{Endler_1977} and often adaptive, but in most cases the selective agent is unknown. For example, in \textit{Drosophila} numerous latitudinal clines exist for traits like thermal tolerance \citep{Hoffmann_etal_2002}, body size (\cite{Coyne_Beecham_1987} and references therein), and life history \citep{Schmidt_etal_2005}. Some \textit{Drosophila} clines have evolved multiple times (\cite{Oakeshott_etal_1982, Huey_etal_2000}, see also \cite{Bradshaw_Holzapfel_2001}) or shifted in response to climate change \citep{Umina_etal_2005}, evincing climatic adaptation. Similarly, plant species exhibit latitudinal clines in traits like flowering time \citep{Stinchcombe_etal_2004}, cyanogenesis \citep{Kooyers_Olsen_2012}, and leaf morphology \citep{Hopkins_etal_2008} that likely related to climatic variation. Despite the fact that latitudinal clines in particular have been studied for a long time, the exact climatic factors (e.g. minimum or maximum temperature, growing season length, strength of biotic interactions) is rarely known. Because many climatic factors vary latitudinally, and which climatic factors vary latitudinally changes over the earth's surface (e.g. coastal vs. continental), dissecting the evolution of latitudinal clines across many species will help biologists identify generalities, such as whether thermal tolerance maxima or seasonal timing is more important \citep{Bradshaw_Holzapfel_2008}. In plants especially, we know little about the prevelance and adaptive significance of clinal variation in fundamental physiological traits like photosynthesis and the impact on plant performance.

@@

% Motivate primary questions using Mimulus
% might add that Anacker and Strauss 2014 showed ecological shifts important in Mim speciation and we would like to know driving forces behind this???
In this study, we address these gaps by asking whether intrinsic or plastic physiological trait variation corresponds with latitude and what climatic factor(s) could plausibly be repsonsible for latitudinal clines in a focal species, \textit{Mimulus cardinalis}. We chose this species because linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, genomic studies under both lab and field conditions. Many classic and contemporary studies of local adaptation use species from genus \textit{Mimulus} because of its natural history, easy propogation, and genetic/genomic resources \citep{Clausen_etal_1940, Hiesey_etal_1971, Bradshaw_Schemske_2003, Wu_etal_2008, Lowry_Willis_2010, Wright_etal_2013}. Yet, there is a conspicuous deficieny of links between local adaptation and physiological mechanisms \citep{Angert_2006, Angert_etal_2008, Wu_etal_2010}. We measured genetic and genotype by environment variation in response to temperature and drought among 16 populations distributed over 10.7\textdegree of latitude. We found a latitudinal cline of intrinsic differences in photosynthesis and growth, but no evidence for variation in plasticity. Interannual variation in precipitaiton and/or precipitation seasonality is associated this axis of variation, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{M. cardinalis}. We place these findings in the context of life history theory and consider future directions in the Discussion.

>>>>

% Local adaptation and clines
Local adaptation within species is ubiquitous; populations generally have higher fitness in their native environment, but perform poorly outside it \citep{Schluter_2000, Hereford_2009}. Local adaptation also frequently leads to clines in both phenotypes and allele frequencies when selection varies over environmental gradients \citep{Huxley_1938, Endler_1977}. Phenotypic differences between populations along a cline most often have a genetic basis and can be studied in a common garden \citep{Turesson_1922, Clausen_etal_1940, Hiesey_etal_1942}. Despite a long history of studying local adaptation and clines, it remains to challegenging to identify exactly which traits are under selection and which differ for nonadaptive reasons. In particular, the role that physiological differences play in local adaptation is poorly understood, despite the fact that physiology is frequently assumed to explain adaptation to the abiotic environment. We need to understand physiological adaptations within species as a baseline for anticipating how organisms will respond to climate change. A related problem is identifying which features of the environment, abiotic factors like soil water availability or biotic interactions, cause spatially varying selective pressures. Here, we examine physiological trait variation and possible selective agents in the perennial herb \textit{Mimulus cardinalis} Douglas ex Benth. (Phrymaceae), a model system for local adaptation studies.

% Intrinsic versus plastic variation
There are two basic approaches one can use to identify candidate traits underlying local adaptation in a common garden. First, if genetically-based trait differences between populations vary clinally with environmental differences, this may point to traits important for local adaptation. Second, genotype by environment interactions could indicate that variation in plasticity mediates local adaptation. We distinguish between these signatures of local adaptation by referring to `intrinsic' and `plastic' trait variation, respectively. There are classic cases of adaptation involving both intrinsic and plastic trait variation. For example, intrinsic differences in critical photoperiod [CITE] and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment. Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genotype by environment interactions cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}. In plants especially, we know little about the prevalance and adaptive significance of variation in fundamental physiological traits like photosynthesis and their impact on plant performance.

% Selective agents
Either instrinsic and/or plastic variation should vary clinally along environmental gradients.\ala{this doesn't make sense given definitions above} Indeed, clines in ecologically important traits are widespread in nature \citep{Endler_1977} and often adaptive, but in most cases the selective agent is unknown. For example, in \textit{Drosophila} numerous latitudinal clines exist for traits like thermal tolerance \citep{Hoffmann_etal_2002}, body size (\cite{Coyne_Beecham_1987} and references therein), and life history \citep{Schmidt_etal_2005}. Some \textit{Drosophila} clines have evolved multiple times (\cite{Oakeshott_etal_1982, Huey_etal_2000}, see also \cite{Bradshaw_Holzapfel_2001}) or shifted in response to climate change \citep{Umina_etal_2005}, evincing climatic adaptation. Similarly, plant species exhibit latitudinal clines in traits like flowering time \citep{Stinchcombe_etal_2004}, cyanogenesis \citep{Kooyers_Olsen_2012}, and leaf morphology \citep{Hopkins_etal_2008} that likely relate to climatic variation.

Despite the fact that latitudinal clines in particular have been studied for a long time, latitude \textit{per se} cannot be a selective agent. Latitude may be strongly correlated with one or two key climatic variables, such as temperature, precipitation, or growing degree-days, such that latitude is an effective proxy for the underlying climatic driver but nonetheless the key climatic variable(s) with stronger relationships to traits are identifiable. Alternatively, latitude may be more strongly related to traits than any single climatic variable for two possible reasons. First, latitude may be correlated with several climatic agents of selection that are individually weak, but add up to a strong latitudinal cline. Alternatively, gene flow among neighboring populations could smooth out local climatic effects, since alleles will experience selection across populations linked by migration. For example, in mountainous regions average temperature at a given latitude varies widely, but in aggregate, a Southern set of populations will experience warmer climate than a Northern one. Thus, any particular Southern population would be warm-adapted, even if it was located in cooler (e.g. high elevation) site. Because many climatic factors vary latitudinally, and which climatic factors vary latitudinally changes over the earth's surface (e.g. coastal vs. continental), dissecting the evolution of latitudinal clines across many species will help biologists identify generalities, such as whether thermal tolerance maxima or seasonal timing is more important \citep{Bradshaw_Holzapfel_2008}. \ala{redo end sentence}

@@

% Motivate primary questions using Mimulus
% might add that Anacker and Strauss 2014 showed ecological shifts important in Mim speciation and we would like to know driving forces behind this???
In this study, we address these gaps by asking whether intrinsic or plastic physiological trait variation corresponds with latitude and what climatic factor(s) could plausibly be repsonsible for latitudinal clines in a focal species, \textit{Mimulus cardinalis}. We chose this species because linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, genomic studies under both lab and field conditions. Many classic and contemporary studies of local adaptation use species from genus \textit{Mimulus} because of its natural history, easy propogation, and genetic/genomic resources \citep{Clausen_etal_1940, Hiesey_etal_1971, Bradshaw_Schemske_2003, Wu_etal_2008, Lowry_Willis_2010, Wright_etal_2013}. Yet, there is a conspicuous deficieny of links between local adaptation and physiological mechanisms \citep{Angert_2006, Angert_etal_2008, Wu_etal_2010}. We measured genetic and genotype by environment variation in response to temperature and drought among 16 populations distributed over 10.7\textdegree of latitude. We found a latitudinal cline of intrinsic differences in photosynthesis and growth, but no evidence for variation in plasticity. Interannual variation in precipitation and/or precipitation seasonality is associated this axis of variation, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{M. cardinalis}. We place these findings in the context of life history theory and consider future directions in the Discussion.
